Lossy compression of DNA sequencing quality data

### 2015-05-14 manuscript/q-compression.tex
bits/pb -> bits/base-call

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile. 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made.

@@

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more memory than the profile encoding methods for
only a $\sim$2--3$\times$ reduction in error. When the rate parameter
of \textsc{q}ual\textsc{c}omp is set to match the profile encoding
methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms
of \textsc{mse}. For the \textit{Rhodobacter sphaeroides} fragment
library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.

>>>>

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile. 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made. We adopt similar ``good''/``bad'' quality value assignments as the preprocessing tool Sickle~\citep{sickle}, which are 40 and 10 respectively.

@@

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more memory than the profile encoding methods for
only a $\sim$2--3$\times$ reduction in error. When the rate parameter
of \textsc{q}ual\textsc{c}omp is set to match the profile encoding
methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms
of \textsc{mse}. For the \textit{Rhodobacter sphaeroides} fragment
library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
(0.099 bits/base-call) has a \textsc{mse} of 17.29, whereas 256-profile
encoding only requires 0.079 bits/base-call and has a \textsc{mse} of 11.85.
